---
title: "Identifying NYC Parks with Vehicle Traffic"
format: 
  html:
    code-fold: true
---
# Research Question
BetaNYC's Research and Data Assistance Request service received the following request from [New Yorkers for Parks](https://www.ny4p.org/):

> *Hi BetaNYC team! Would you be able to help us with a list of NYC Parks that allow personal vehicles to drive through them, prioritizing parks where pedestrians' and cyclists' experience in the park is shaped by vehicles (vs. Central Park transverse example, which is separated from people)? [...] Not sure how to go about this except by looking at a map, park by park and would love BetaNYC's support.*

Only two New York City parks have banned vehicle traffic: Central Park in Manhattan and Prospect Park in Brooklyn[^1]. It appears there is no official list of parks that *do* permit vehicles in parks. A tentative list of parks that do may be derived from the assumption that **any park that contains a road segment suitable for vehicle traffic, with the exception of Central Park and Prospect Park, allows vehicle traffic.**

```{r}
#| label: setup
#| include: true
#| warning: false
#| message: false

# Notebook setup

# Use quarto render to generate a static site in /docs
# Set github pages to deploy from main branch, /docs

# Insure packages are installed by checking required packages against installed packages
reqd_pkgs <- c("flextable",
               "sf",
               "tidyverse",
               "mapgl",
               "htmltools",
               "officer",
               "forcats",
               "socratadata",
               "mapgl",
               "mapview",
               "DBI",
               "RPostgres")

install.packages(
  setdiff(
    reqd_pkgs,
    rownames(installed.packages())
  ),
  repos = "https://cloud.r-project.org"
)

lapply(reqd_pkgs, library, character.only = TRUE) |> 
  invisible()

# If needed, ensure that the Quarto render and RStudio read the same environment vars
readRenviron(".Renviron")

# Start Postgres via Docker, wait for healthcheck
system("docker compose up -d --wait")

# Set up connection from env
con <- dbConnect(RPostgres::Postgres())

# Enable PostGIS
invisible(dbExecute(con, "CREATE EXTENSION IF NOT EXISTS postgis"))

# Get cil_table()
source("cil_table.R")
```

# Available Data

Two data sources can be used to derive the list:

1. [Parks Properties](https://data.cityofnewyork.us/Recreation/Parks-Properties/enfh-gkve/about_data)
2. [NYC Street Centerline](https://data.cityofnewyork.us/City-Government/Centerline/inkn-q76z/about_data)

## Parks Properties

The Parks Properties dataset description carries the caveat:

> *This dataset identifies property managed partially or solely by NYC Parks. This data has been produced in whole or part using secondary data. Data accuracy is limited by the scale and accuracy of the original sources. Site-specific conditions should be field-verified. *

At least two ambiguities are described above. One is the official-unofficial nature of the data. It's provided to NYC Open Data by the Department of Parks, but it's comprised of secondary data (at least in part, or maybe in whole -- another ambiguity. Does the Department of Parks not know what properties are under their jurisdiction?). The other is that it's unclear what types of properties are included in the dataset. Are they all parks? A mixture of parks and something else? We can take a look at the unique combinations of `class`, `subcategory`, and `typecategory`:

```{r}
invisible(
  property_types <- soc_read(
    "https://data.cityofnewyork.us/Recreation/Parks-Properties/enfh-gkve.json",
    query = soc_query(
      select = "distinct class, subcategory, typecategory"
    )
  )
)

cil_table(property_types)
```

`PARK` is the only `class` of property in the dataset. Based on the combinations of `subcategory` and `typecategory`, it may be possible to determine which properties may be excluded from our list of parks, but for now we'll consider each feature in the dataset to be a park.

## NYC Street Centerline

The centerline dataset is described as:

> *The NYC Street Centerline (CSCL) is a road-bed representation of New York City streets containing address ranges and other information such as traffic directions, road types, segment types.*

This brief description implies that this dataset is entirely composed of streets, i.e. surfaces for vehicles to travel. However, taking a look at the `rw_type` reveals that not to be the case:

```{r}
street_types <- property_types <- soc_read(
  "https://data.cityofnewyork.us/City-Government/Centerline/inkn-q76z.json",
  query = soc_query(
    select = "distinct rw_type"
  )
)

street_types <- street_types |> 
  mutate(`Roadway Type` = case_when(
    rw_type == 1 ~ "Street",
    rw_type == 2 ~ "Highway",
    rw_type == 3 ~ "Bridge",
    rw_type == 4 ~ "Tunnel",
    rw_type == 5 ~ "Boardwalk",
    rw_type == 6 ~ "Path/Trail",
    rw_type == 7 ~ "StepStreet",
    rw_type == 8 ~ "Driveway",
    rw_type == 9 ~ "Ramp",
    rw_type == 10 ~ "Alley",
    rw_type == 11 ~ "Unknown",
    rw_type == 12 ~ "Non-Physical Street Segment",
    rw_type == 13 ~ "U Turn",
    rw_type == 14 ~ "Ferry Route",
  ))

street_types |>
  arrange(rw_type) |>
  mutate(rw_type = as.character(rw_type)) |>
  cil_table(widths = c(15, 85),
            unit = "%")
```

We'll want to exclude segments where `rw_type` is `6` to keep paths/trails from counting as a site for vehicular traffic. It's possible that other roadway types should be excluded, but for now, only `Path/Trail` will be. 

# Considerations

One consideration for this analysis are large parks spanning city streets. Take, for example, McCarren Park in Greenpoint, Brooklyn, which is comprised of five separate areas:

```{r}
mccarren <- soc_read(
  "https://data.cityofnewyork.us/Recreation/Parks-Properties/enfh-gkve.json",
  query = soc_query(
    where = "signname = 'McCarren Park'" 
  )
)

maplibre(
  bounds = mccarren |> st_buffer(dist = 150),
  style = carto_style("positron")
) |> 
  add_line_layer(
    id = "mccarren-park",
    source = mccarren,
    line_color = "#3258a3",
    line_opacity = 0.85,
    line_width = 2
  )
```

If our consideration of a park that permits vehicle traffic is strict, McCarren Park will not qualify unless there are road segments *within* any of the separate areas. In order to capture the broader definition used by New Yorkers for Parks of "parks where pedestrians' and cyclists' experience in the park is shaped by vehicles", we can use the **convex hull** operation, which is analogous to the area enclosed by wrapping a rubber band around multi-part areas:

```{r}
mcc_ch <- mccarren |> st_drop_geometry()

mcc_ch$geometry <- st_convex_hull(mccarren$multipolygon)

st_geometry(mcc_ch) <- mcc_ch$geometry


maplibre(
  bounds = mccarren |> st_buffer(dist = 150),
  style = carto_style("positron")
) |> 
  add_line_layer(
    id = "mccarren-park",
    source = mcc_ch,
    line_color = "#3258a3",
    line_opacity = 0.85,
    line_width = 2
  )

```

Care must be taken with this approach, though, as the operation may introduce roads into the resulting area that as an artifact. See above along the southwest border of the park where North 12th Street now falls within the new park area, where it didn't overlap before. 

**All to say: This more inclusive definition of park area by necessity introduces false positives into the list. Manual review is required**

# Methodology

1. Fetch road segments which are not `Path/Trail`
2. Fetch parks properties
3. Get convex hull of property boundaries to enclose disjoint parks
4. Use spatial inner join with `ST_INTERSECTS` spatial predicate to identify parks with intersecting street segments


```{r, cache=TRUE, message=FALSE, warning=FALSE}
centerline <- soc_read(
  "https://data.cityofnewyork.us/City-Government/Centerline/inkn-q76z.json",
  query = soc_query(
    where = "rw_type != 6"
  )
) |> st_make_valid()

parks <- soc_read(
  "https://data.cityofnewyork.us/Recreation/Parks-Properties/enfh-gkve.json"
) |> 
  st_make_valid()

# rename geometry cols
st_geometry(centerline) <- "geom"
st_geometry(parks) <- "geom"
```

```{r, message=FALSE, warning=FALSE}
# Transform to state plane (ft), write to db
st_write(st_transform(centerline, 2263), con, "centerline", delete_layer = TRUE, quiet = TRUE)
st_write(st_transform(parks, 2263), con, "parks", delete_layer = TRUE, quiet = TRUE)

# Compute convex hulls into new table for fast query (rather than as CTE)
invisible(dbExecute(con, "DROP TABLE IF EXISTS parks_ch"))
invisible(
  dbExecute(con, "
    CREATE TABLE parks_ch AS
    SELECT gispropnum, signname, ST_ConvexHull(ST_MakeValid(geom)) AS geom
    FROM parks")
)
invisible(dbExecute(con, "CREATE INDEX ON parks_ch USING GIST (geom)"))
invisible(dbExecute(con, "ANALYZE parks_ch"))

parks_with_roads_all <- dbGetQuery(con, "
  SELECT p.gispropnum, p.signname, c.physicalid, c.rw_type
  FROM parks_ch p
  JOIN centerline c ON ST_Intersects(p.geom, c.geom)
  ORDER BY p.gispropnum
")

dbDisconnect(con)

park_list <- parks |> 
  filter(gispropnum %in% parks_with_roads_all$gispropnum)

road_list <- centerline |> 
  filter(physicalid %in% parks_with_roads_all$physicalid)


```

With this approach, `r nrow(park_list)` of `r scales::comma(nrow(parks))` NYC parks are found to have overlapping roads. They are displayed below:

```{r}
maplibre(
  bounds = park_list |> st_buffer(500),
  style = carto_style("positron")
) |>
  add_source(
    id = "parks-src",
    data = select(park_list, gispropnum, signname),  # only the columns the map needs
    tolerance = 0                                    # no simplification at any zoom
  ) |>
  # Transparent fill layer for tooltip 
  add_fill_layer(
    id = "parks-hover",
    source = "parks-src",
    fill_color = "#3258a3",
    fill_opacity = 0,
    tooltip = "signname"
  ) |>
  add_line_layer(
    id = "parks-with-roads",
    source = "parks-src",
    line_color = "#3258a3",
    line_opacity = 0.85
  )
```

As noted earlier, the approach used to identify roads splitting up multi-part parks **necessarily introduces false positives**. The list of `r nrow(park_list)` needs to be manually reviewed. 

Consider the case of Hoffman Park in Queens: The convex hull boundary (red) overlaps with the ramps on and off of the LIE from Woodhaven Boulevard, whereas the actual boundary (blue) does not:

```{r, message=FALSE}
hoffman_park_original <- parks |> 
  filter(name311 == "Hoffman Park") |> 
  select(gispropnum, name311)

hoffman_park_ch <- hoffman_park_original |> 
  st_drop_geometry()

hoffman_park_ch$geom <- st_convex_hull(hoffman_park_original$geom)

st_geometry(hoffman_park_ch) <- "geom"

dir.create("output", showWarnings = FALSE)

segment_counts <- count(parks_with_roads_all, gispropnum, name = "road_segments")

park_list |>
  st_drop_geometry() |>
  select(gispropnum, signname, borough, typecategory, subcategory, acres) |>
  left_join(segment_counts, by = "gispropnum") |>
  arrange(borough, signname) |>
  write_csv("output/parks_with_roads.csv", na = "")

invisible(
  park_list |>
    select(gispropnum, signname, borough, typecategory, subcategory, acres) |>
    left_join(segment_counts, by = "gispropnum") |>
    arrange(borough, signname) |>
    st_write("output/parks_with_roads.geojson",
             delete_dsn = TRUE,
             quiet = TRUE)
)

maplibre(
  bounds = hoffman_park_original |> st_buffer(10),
  style = carto_style("positron")
) |>  
  add_line_layer(
    id = "hoffman-original",
    source = hoffman_park_original,
    line_color = "#3258a3",
    line_opacity = 0.85,
    line_width = 2
  ) |> 
  add_line_layer(
    id = "hoffman-ch",
    source = hoffman_park_ch,
    line_color = "#973835",
    line_opacity = 0.85,
    line_width = 2
  )
```

Review the tentative list of `r nrow(park_list)` parks: [parks_with_roads.csv](output/parks_with_roads.csv){download="parks_with_roads.csv"}

Or download the spatial file here: [parks_with_roads.geojson](output/parks_with_roads.geojson){download="parks_with_roads.geojson"}

[^1]: [NYC Parks: Car-Free Park Hours](https://www.nycgovparks.org/highlights/car-free-park-hours)